Resolving a paradox of measurement uncertainty in ecological management

### 2018-08-17 manuscripts/manuscript.tex
here we go

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The importance of measurement uncertainty in ecological management},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{The importance of measurement uncertainty in ecological management}
    \author[]{Authors removed during peer review}

  \begin{abstract}
  Ecological management and decision-making typically focus on uncertainty
  about the future, but surprisingly little is known about how to account
  for uncertainty of the present: that is, the realities of having only
  partial or imperfect measurements. Our two primary paradigms for
  handling decisions under uncertainty -- the precautionary principle and
  optimal control -- have so far given entirely contradictory results.
  This paradox is best illustrated in the example of fisheries management,
  where many ideas that guide thinking about ecological decision making
  were first developed. We find that simplistic optimal control approaches
  have repeatedly concluded that a manager should increase catch quotas
  when faced with greater uncertainty about the fish biomass. Current best
  practices take a more precautionary approach, decreasing catch quotas by
  a fixed amount to account for uncertainty. Using comparisons to both
  simulated and historical catch data, we find that neither approach is
  sufficient to avoid stock collapses under moderate observational
  uncertainty. Using partially observed Markov decision process (POMDP)
  methods, we demonstrate how this paradox arises from flaws in the
  standard theory, which contributes to over-exploitation of fisheries and
  increased probability of economic and ecological collapse. In contrast,
  we find POMDP-based management avoids such over-exploitation while also
  generating higher economic value. These results have significant
  implications for how we handle uncertainty in both fisheries and
  ecological management more generally.
  \end{abstract}
   \begin{keyword} Measurement uncertainty POMDP Decision theory Fisheries Conservation\end{keyword}
 \end{frontmatter}

@@

The challenge of making decisions under uncertainty is obviously not
unique to fisheries, and the existing methods we critique here can be
found in a wide range of population and ecosystem management challenges,
including fire ecology (Richards et al. 1999), invasive species (Shea
1998; Blackwood et al. 2010), disease outbreaks (Shea et al. 2014; Li et
al. 2017), and protecting biodiversity (Dee et al. 2017; Iacona et al.
2017). Yet fisheries conservation and management has long been both a
crucible and proving ground for the theory of ecological management more
generally, including topics such as adaptive management (Walters and
Hilborn 1978), ecosystem-based management (Levin and Lubchenco 2008) and
resilience thinking (Holling 1973; May 1977), while also giving rise to
the sub-discipline of resource economics (Gordon and Press 1954;
Schaefer 1954; Beverton and Holt 1957), thanks to its global relevance,
long history, and readily available data. Despite the generality of the
models and issue of measurement uncertainty considered here, any
application must be grounded in both the history and context of the
particular decision problem. In light of this, we begin with a brief
background on existing decision theory approaches common in fisheries
management.

In this paper, we examine three strategies to managing a natural
population, such as a fishery, under such uncertainty. This problem has
a long history in both theory and practice, where seemingly innoccous
simplifying assumptions have led to a deep paradox in how managers
should respond to uncertainty. The precautionary principle (Kriebel et
al. 2001) would seem to suggest that the greater the uncertainty in our
models, the more cautious we should be in management policies. Yet a
well-known mathematical proof by Reed (1979) established that under
quite general conditions, the optimal strategy under uncertainty is no
different than under no uncertainty at all. Attempts to resolve this
have since only deepened this paradox, arguing that adding additional
sources of uncertainty, such as imperfect measurements, should lead to
optimal harvest that increases with greater uncertainty (e.g. Clark and
Kirkwood 1986; Sethi et al. 2005). Meanwhile, much of actual practice
has (perhaps fortunately, as we shall see) ignored these results in
favor of a more heuristic application of the precautionary principle.
Here, we show that this paradox can be resolved by uncovering previous
shortcuts and tackling the optimal management problem head-on, which we
can solve thanks to the computational efficiency of sophisticated
algorithms developed more recently in the field of robotics. This
approach allows us to reject this paradox largely in favor of the
precautionary principle. Further, our comparisons show that widely
recognized existing approaches in both theory and practice are not by
themselves sufficient to manage a population under substantial
uncertainty over the long term, but our recently borrowed approach from
robotics literature can under identical conditions sustain both higher
biomass and higher economic yields than those approaches. This
conclusion has important implications not only for fisheries but
eocsystem more broadly, underscoring the danger of not correctly
accounting for measurement uncertainty when making decisions, and
further to the significant opportunity to learn from fields such as
robotics when faced with complex and uncertain decisions.

@@

We consider the management problem of setting catch quotas for a marine
fishery in the face of imperfect information about the current stock
size and uncertainty about future recruitment. We seek to determine the
sequence of actions \(h_t\) for \(t \in [1, ... \infty]\) that maximize
the net present value (discounted sum of all future profits) of the
fishery. We will denote the discount rate \(\gamma\). For simplicity we
will once again follow classic theory and assume a fixed price for fish,
(equivalently, measuring our value in units of discounted fish rather
than discounted dollars), though again this assumption is easy to relax.
Each year, the manager also obtains an estimate \(y_t\) of the true
\(x_t\) stock size subject to some measurement uncertainty \(\xi_t\)

@@

For simplicity we will assume measurement uncertainty is normally
distributed with standard deviation \(\sigma_m\),
\(\xi \sim \mathcal{N}(1, \sigma_m)\). Given an estimated population
model along with this (uncertain) measurement of the stock size, the
manager must choose the set of a harvest quotas \(\lbrace h_t \rbrace\)
for \(t \in [0, 1, \infty]\) that maximizes expected long term utility
of the stock:

@@

where the \(\min\) function ensures no additional reward for attempting
to harvest more than the entire stock. Existing optimal control
techniques (Reed (1979)) prove these results are not dependent on the
particular details of this utility. Using this form allows a more direct
comparison to approaches such as maximum sustainable yield (MSY), that
do not include an economic model but merely maximize long-term catch. In
principle, utility could also reflect ecosystem services or other
intrinsic values for fish left in the sea and not merely the market for
consumption (Halpern et al. 2013).

@@

Maximum Sustainable Yield (MSY) remains the basis of international law
(including the UN, IWC, IATTC, ICCAT, ICNAF; Mace (2001)) and a familiar
standard of management in terrestrial ecosystems as well as aquatic
(Clark 1990). Critics have for some time observed the limitations of
harvesting at MSY in face of uncertainty in stock sizes and population
dynamics (Larkin 1977; Botsford et al. 1997). Many US fisheries reflect
this uncertainty through a series of adjustments that effectively reduce
the target fishing mortality level to reflect this uncertainty.
Typically, a stock assessment model provides a best estimate of
\(B_{MSY}\) and corresponding mortality \(F_{MSY}\) is used to define
the stock Over-Fishing Limit, (OFL). Based on this, a somewhat lower
level is set as the Allowable Biological Catch (ABC), reflecting
uncertainty in the stock assessment. To reflect possible uncertainty
between reported and actual catch, the ABC is reduced somewhat further
to define the Total Allowable Catch, (TAC), which forms the basic unit
of management for many such fisheries. To reflect this process, our
analysis will also consider policies in which the harvest quota is set
at 80\% of the level expected under the MSY policy:
\(H_{TAC}(t) = 0.8 \cdot F_{MSY} \cdot B_t\), for a biomass estimated at
\(B(t)\) in year \(t\). To distinguish this approach from MSY, we will
refer to this approach as a TAC policy. This more closely represents a
heuristic (e.g. Hilborn 2010) or resilience-based approach (sensu
Fischer et al. 2009) than an optimization-based policy. Importantly,
this approach shares the fundamentally stationary assumptions of an MSY
policy by defining a constant mortality rather than a dynamic policy.
Thus, despite being more cautious overall and generating lower economic
yield than a constant escapement policy, TAC policies continue to
harvest at non-zero rates even if a stock falls below \(B_{MSY}\), while
the constant escapement policy does not.

It is important to note that actual management strategies are quite
diverse. Our focus on comparisons TAC, MSY, and Constant Escapement (CE,
also refereed to as MEY, see below) reflects only the subset of current
practices. Another increasingly common approach to address uncertainty
in both process and models is known as Management Strategy Evaluation
(MSE, Smith 1994; Dorner et al. 2009; Bunnefeld et al. 2011; Punt et al.
2016). MSE performs forward simulations under a suite of pre-selected
candidate strategies to determine which strategy achieves the best
expected outcomes. Crucially, by limiting the analysis to a suite of
candidate strategies over a finite number of replicate simulations, this
approach is able to accommodate substantially greater complexity and
uncertainty than has been possible with more formal optimal control
approaches we discuss below. Optimal control solutions rely on dynamic
programming to consider all possible strategies (all possible sequences
of harvest under all possible realizations of stock dynamics). As such,
MSE can be viewed as an approximation to the optimal control solutions
considered below, a notion that has been formalized in the literature on
approximate dynamic programming (Nicol and Chadès 2011).

@@

We consider the average fish biomass across 100 replicate simulations of
the same stock dynamics (Eq 4) under three different management
strategies: constant escapement (CE), total allowable catch (TAC; equal
to 80\% MSY), and partially observed Markov decision process (POMDP)
management. Each suite of simulations is considered under three
sequentially higher measurement error regimes, from \(\sigma_m = 0\) (no
measurement errors), \(\sigma_m = 0.1\) (low error), and
\(\sigma_m = 0.15\) (moderate error), as indicated, while stochastic
recruitment (environmental noise) is set to a moderate
\(\sigma_g = 0.15\). Empirical estimates of measurement error in stock
sizes vary widely, with estimates often dependent upon methodological
assumptions. Early estimates have put possible error as high as 50\%
(Clark and Kirkwood 1986) and more recent analyses have suggested an
average of closer to 36\%, though with cases that may be much higher
(Ralston et al. 2011). Consequently, range between 0 to 15\% provides a
probably conservative estimate of typical unceratinty ranges, which we
show is nevertheless large enough to drive substantially negative
ecological and economic outcomes under management strategies that do not
appropriately account for that error over the long run.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Resolving a paradox of measurement uncertainty in ecological management},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Resolving a paradox of measurement uncertainty in ecological management}
    \author[]{Authors removed during peer review}

  \begin{abstract}
  Ecological management and decision-making typically focus on uncertainty
  about the future, but surprisingly little is known about how to account
  for uncertainty of the present: that is, the realities of having only
  partial or imperfect measurements. Our primary paradigms for handling
  decisions under uncertainty -- the precautionary principle and optimal
  control -- have so far given contradictory results. This paradox is best
  illustrated in the example of fisheries management, where many ideas
  that guide thinking about ecological decision making were first
  developed. We find that simplistic optimal control approaches have
  repeatedly concluded that a manager should increase catch quotas when
  faced with greater uncertainty about the fish biomass. Current best
  practices take a more precautionary approach, decreasing catch quotas by
  a fixed amount to account for uncertainty. Using comparisons to both
  simulated and historical catch data, we find that neither approach is
  sufficient to avoid stock collapses under moderate observational
  uncertainty. Using partially observed Markov decision process (POMDP)
  methods, we demonstrate how this paradox arises from flaws in the
  standard theory, which contributes to over-exploitation of fisheries and
  increased probability of economic and ecological collapse. In contrast,
  we find POMDP-based management avoids such over-exploitation while also
  generating higher economic value. These results have significant
  implications for how we handle uncertainty in both fisheries and
  ecological management more generally.
  \end{abstract}
   \begin{keyword} POMDP, measurement uncertainty, decision theory, fisheries, conservation\end{keyword}
 \end{frontmatter}

@@

The challenge of making decisions under uncertainty is obviously not
unique to fisheries. Existing methods we optimal control methods which
similarly assume perfect measurements can be found in a wide range of
population and ecosystem management challenges, including fire ecology
(Richards et al. 1999), invasive species (Shea 1998; Blackwood et al.
2010), disease outbreaks (Shea et al. 2014; Li et al. 2017), and
protecting biodiversity (Dee et al. 2017; Iacona et al. 2017), and in
common textbooks and reviews (Mangel 1985; Clark 1990; Marescot et al.
2013). Yet fisheries conservation and management has long been both a
crucible and proving ground for the theory of ecological management more
generally, including topics such as adaptive management (Walters and
Hilborn 1978), ecosystem-based management (Levin and Lubchenco 2008) and
resilience thinking (Holling 1973; May 1977), while also giving rise to
the sub-discipline of resource economics (Gordon and Press 1954;
Schaefer 1954; Beverton and Holt 1957), thanks to its global relevance,
long history, and readily available data. Despite the generality of the
models and issue of measurement uncertainty considered here, any
application must be grounded in both the history and context of the
particular decision problem. In light of this, we begin with a brief
background on existing decision theory approaches common in fisheries
management.

In this paper, we examine three strategies to managing a natural
population, such as a fishery, under such uncertainty. This problem has
a long history in both theory and practice, where seemingly innocuous
simplifying assumptions have led to a deep paradox in how managers
should respond to uncertainty. The precautionary principle (Kriebel et
al. 2001) would seem to suggest that the greater the uncertainty in our
models, the more cautious we should be in management policies. Yet a
well-known mathematical proof by Reed (1979) established that under
quite general conditions, the optimal strategy under uncertainty is no
different than under no uncertainty at all. Attempts to resolve this
have since only deepened this paradox, arguing that adding additional
sources of uncertainty, such as imperfect measurements, should lead to
optimal harvest that increases with greater uncertainty (e.g. Clark and
Kirkwood 1986; Sethi et al. 2005). Meanwhile, much of actual practice
has (perhaps fortunately, as we shall see) ignored these results in
favor of a more heuristic application of the precautionary principle.
Here, we show that this paradox can be resolved by uncovering previous
shortcuts and tackling the optimal management problem head-on, which we
can solve thanks to the computational efficiency of sophisticated
algorithms developed more recently in the field of robotics. This
approach allows us to reject this paradox largely in favor of the
precautionary principle. Further, our comparisons show that widely
recognized existing approaches in both theory and practice are not by
themselves sufficient to manage a population under substantial
uncertainty over the long term, but our recently borrowed approach from
robotics literature can under identical conditions sustain both higher
biomass and higher economic yields than those approaches. This
conclusion has important implications not only for fisheries but
ecosystem more broadly: underscoring the danger of not correctly
accounting for measurement uncertainty when making decisions, and
further to the significant opportunity to learn from fields such as
robotics when faced with complex and uncertain decisions.

@@

We consider the management problem of setting catch quotas for a marine
fishery in the face of imperfect information about the current stock
size and uncertainty about future recruitment. We seek to determine the
sequence of actions \(h_t\) for \(t \in [1, ... \infty]\) that maximize
the net present value (discounted sum of all future profits) of the
fishery. We will denote the discount rate \(\gamma\). For simplicity we
will once again follow classic theory and assume a fixed price for fish,
(equivalently, measuring our value in units of discounted fish rather
than discounted dollars). Each year, the manager also obtains an
estimate \(y_t\) of the true \(x_t\) stock size subject to some
measurement uncertainty \(\xi_t\)

@@

For simplicity, we will assume measurement uncertainty is normally
distributed with standard deviation \(\sigma_m\),
\(\xi \sim \mathcal{N}(1, \sigma_m)\). Given an estimated population
model along with this (uncertain) measurement of the stock size, the
manager must choose the set of a harvest quotas \(\lbrace h_t \rbrace\)
for \(t \in [0, 1, \infty]\) that maximizes expected long term utility
of the stock:

@@

where the \(\min\) function ensures no additional reward for attempting
to harvest more than the entire stock. In economic terms, the
corresponds to a fixed price per fish caught, with no proportional cost
on effort required to do so, but the results do not depend on this
assumption, as existing analyses have previously established (Reed
1979). Using this form allows a more direct comparison to approaches
such as maximum sustainable yield (MSY) which do not include an economic
model but merely maximize long-term catch. In principle, utility could
also reflect ecosystem services or other intrinsic values for fish left
in the sea and not merely the market for consumption (Halpern et al.
2013).

@@

Maximum Sustainable Yield (MSY) remains the basis of international law
(including the UN, IWC, IATTC, ICCAT, ICNAF; Mace (2001)) and a familiar
standard of management in terrestrial ecosystems as well as aquatic
(Clark 1990). Critics have for some time observed the limitations of
harvesting at MSY in face of uncertainty in stock sizes and population
dynamics (Larkin 1977; Botsford et al. 1997). Many US fisheries reflect
this uncertainty through a series of adjustments that effectively reduce
the target fishing mortality level to reflect this uncertainty.
Typically, a stock assessment model provides a best estimate of
\(B_{MSY}\) and corresponding mortality \(F_{MSY}\) is used to define
the stock Over-Fishing Limit, (OFL). Based on this, a somewhat lower
level is set as the Allowable Biological Catch (ABC), reflecting
uncertainty in the stock assessment. To reflect possible uncertainty
between reported and actual catch, the ABC may be reduced further to
define the Total Allowable Catch, (TAC), which forms the basic unit of
management for many such fisheries. To reflect this process, our
analysis will also consider policies in which the harvest quota is set
at 80\% of the level expected under the MSY policy:
\(H_{TAC}(t) = 0.8 \cdot F_{MSY} \cdot B_t\), for a biomass estimated at
\(B(t)\) in year \(t\). To distinguish this approach from MSY, we will
refer to this approach as a TAC policy. This more closely represents a
heuristic (e.g. Hilborn 2010) or resilience-based approach (sensu
Fischer et al. 2009) than an optimization-based policy. Importantly,
this approach shares the fundamentally stationary assumptions of an MSY
policy by defining a constant mortality rather than a dynamic policy.
Thus, despite being more cautious overall and generating lower economic
yield than a constant escapement policy, TAC policies continue to
harvest at non-zero rates even if a stock falls below \(B_{MSY}\), while
the constant escapement policy does not.

It is important to note that actual management strategies are quite
diverse. Our focus on comparisons TAC, MSY, and Constant Escapement (CE)
reflects only the subset of current practices. Another increasingly
common approach to address uncertainty in both process and models is
known as Management Strategy Evaluation (MSE, Smith 1994; Dorner et al.
2009; Bunnefeld et al. 2011; Punt et al. 2016). MSE performs forward
simulations under a suite of pre-selected candidate strategies to
determine which strategy achieves the best expected outcomes. Crucially,
by limiting the analysis to a suite of candidate strategies over a
finite number of replicate simulations, this approach is able to
accommodate substantially greater complexity and uncertainty than has
been possible with more formal optimal control approaches we discuss
below. Optimal control solutions rely on dynamic programming to consider
all possible strategies (all possible sequences of harvest under all
possible realizations of stock dynamics). As such, MSE can be viewed as
an approximation to the optimal control solutions considered below, a
notion that has been formalized in the literature on approximate dynamic
programming (Nicol and Chadès 2011). Future work may be able to leverage
the forward-simulation approach used in MSE to extend the POMDP methods
introduced here to yet more complex dynamics.

@@

We consider the average fish biomass across 100 replicate simulations of
the same stock dynamics (Eq 4) under three different management
strategies: constant escapement (CE), total allowable catch (TAC; equal
to 80\% MSY), and partially observed Markov decision process (POMDP)
management. Each suite of simulations is considered under three
sequentially higher measurement error regimes, from \(\sigma_m = 0\) (no
measurement errors), \(\sigma_m = 0.1\) (low error), and
\(\sigma_m = 0.15\) (moderate error), as indicated, while stochastic
recruitment (environmental noise) is set to a moderate
\(\sigma_g = 0.15\). Empirical estimates of measurement error in stock
sizes vary widely, with estimates often dependent upon methodological
assumptions. Early estimates have put possible error as high as 50\%
(Clark and Kirkwood 1986) and more recent analyses have suggested an
average of closer to 36\%, though with cases that may be much higher
(Ralston et al. 2011). Consequently, range between 0 to 15\% provides a
probably conservative estimate of typical uncertainty ranges, which we
show is nevertheless large enough to drive substantially negative
ecological and economic outcomes under management strategies that do not
appropriately account for that error over the long run.
